Review on precise spiking motifs in neurobiological and neuromorphic data

### 2022-10-04 manuscript.tex
Update on Overleaf.

% !TeX TS-program = pdfLaTeX
% !TeX encoding = UTF-8 Unicode
% !TeX spellcheck = en-US
% !BIB TS-program = bibtex
% -*- coding: UTF-8; -*-
% vim: set fenc=utf-8
% https://github.com/SpikeAI/2022-09_UltraFastCat/blob/main/Readme.md
% https://www.overleaf.com/project/630dcf8012267a5bb40967dd
% https://framateam.org/spikeai/channels/ultrafastcat
% https://www.zotero.org/groups/4560566/ultrafastcat
% https://arxiv.org/abs/2205.03635
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
%A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.
Why do neurons communicate through spikes? By definition, a spike, or action potential, is a all-or-none event ---it can occur or not without further details--- at asynchronous timings, i.e.~it can occur at any continuous time - differentially to the discretized timing classically used in digital processing. In the living world, neural systems almost systematically use this so-called event-based representation, though we do not yet have a clear idea why. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging challenge in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal challenge in the age of the digital economy and of global warming.

The response of a biological neuron depends largely on the precise timing of the sequence of presynaptic spikes as they reach the basal dendritic tree. This \emph{event-based representation} present in the neuronal code is essential in understanding information processing and yet, most neuronal models do not take advantage of this minute temporal dimension. Our goal here is to bring an interdisciplinary perspective on the computational advantage of time series representations for the brain and for information processing machines. In particular, we will focus on the hypothesis that there exists in an assembly of neurons a representation based on a set of motifs of different relative spike times. Here, we will review current litterature on the detection of such motifs in generic raster plots. It is work in progress, where anybody interested can \emph{openly} join.

This hypothesis is directly inspired by neurobiological observations in the hippocampus, and it expands the capabilities of analog representations based on the firing rate by considering a representation based on repetitions of these motifs at precise times of occurrence. A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data. We will first review some biological and theoretical evidence in neural information processing. We will then present some models for the detection of such motifs in arbitrary raster plots, synthetic, biological or artificial (notably from event-based cameras). In particular, we will discuss models which exploit the variety of synaptic delays on the dendritic tree. Then, we will try to outline some possible strategies for learning these patterns and finally discuss possible perspectives.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\setcounter{section}{-1} %% Remove this when starting to work on the template.
\section{How to Use this Template}

The template details the sections that can be used in a manuscript. Note that the order and names of article sections may differ from the requirements of the journal (e.g., the positioning of the Materials and Methods section). Please check the instructions on the authors' page of the journal to verify the correct order and names. For any questions, please contact the editorial office of the journal or support@mdpi.com. For LaTeX-related questions please contact latex@mdpi.com.%\endnote{This is an endnote.} % To use endnotes, please un-comment \printendnotes below (before References). Only journal Laws uses \footnote.

% The order of the section titles is: Introduction, Materials and Methods, Results, Discussion, Conclusions for these journals: aerospace,algorithms,antibodies,antioxidants,atmosphere,axioms,biomedicines,carbon,crystals,designs,diagnostics,environments,fermentation,fluids,forests,fractalfract,informatics,information,inventions,jfmk,jrfm,lubricants,neonatalscreening,neuroglia,particles,pharmaceutics,polymers,processes,technologies,viruses,vision

\section{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Patents}

This section is not mandatory, but may be added if there are patents resulting from the work reported in this manuscript.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

@@

\dataavailability{In this section, please provide details regarding where data supporting reported results can be found, including links to publicly archived datasets analyzed or generated during the study. Please refer to suggested Data Availability Statements in section ``MDPI Research Data Policies'' at \url{https://www.mdpi.com/ethics}. If the study did not report any data, you might add ``Not applicable'' here.}

>>>>

% !TeX TS-program = pdfLaTeX
% !TeX encoding = UTF-8 Unicode
% !TeX spellcheck = en-US
% !BIB TS-program = bibtex
% -*- coding: UTF-8; -*-
% vim: set fenc=utf-8
% https://www.overleaf.com/project/633bf90497b5b325a3ff692e
% https://framateam.org/spikeai/channels/review_polychrony
% https://www.zotero.org/groups/4562620/polychronies
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%  LaTeX support: latex@mdpi.com 
%  For support, please attach all files needed for compiling as well as the log file, and specify your operating system, LaTeX version, and LaTeX editor.

@@

% Abstract (Do not insert blank lines, i.e. \\) 
\abstract{
%A single paragraph of about 200 words maximum. For research articles, abstracts should give a pertinent overview of the work. We strongly encourage authors to use the following style of structured abstracts, but without headings: (1) Background: place the question addressed in a broad context and highlight the purpose of the study; (2) Methods: describe briefly the main methods or treatments applied; (3) Results: summarize the article's main findings; (4) Conclusions: indicate the main conclusions or interpretations. The abstract should be an objective representation of the article, it must not contain results which are not presented and substantiated in the main text and should not exaggerate the main conclusions.
Why do neurons communicate through spikes? By definition, a spike, or action potential, is a all-or-none event ---it can occur or not without further details--- at asynchronous timings, i.e.~it can occur at any continuous time - differentially to the discretized timing classically used in digital processing. In the living world, neural systems almost systematically use this so-called event-based representation, though we do not yet have a clear idea why. A better understanding of this phenomenon remains a fundamental challenge in neurobiology in order to better interpret the masses of recorded data. It is also an emerging challenge in computer science to allow the efficient exploitation of a new class of sensors and event-based computers, called neuromorphic, which could allow significant gains in computing time and energy consumption ---a major societal challenge in the age of the digital economy and of global warming.
%
The response of a biological neuron depends largely on the precise timing of the sequence of presynaptic spikes as they reach the basal dendritic tree. This \emph{event-based representation} present in the neuronal code is essential in understanding information processing and yet, most neuronal models do not take advantage of this minute temporal dimension. Our goal here is to bring an interdisciplinary perspective on the computational advantage of time series representations for the brain and for information processing machines. In particular, we will focus on the hypothesis that there exists in an assembly of neurons a representation based on a set of motifs of different relative spike times. Here, we will review current litterature on the detection of such motifs in generic raster plots. It is work in progress, where anybody interested can \emph{openly} join.
%
This hypothesis is directly inspired by neurobiological observations in the hippocampus, and it expands the capabilities of analog representations based on the firing rate by considering a representation based on repetitions of these motifs at precise times of occurrence. A mathematical formalization would be particularly well suited to neuromorphic computing, and would allow for the supervised or self-supervised learning of such motifs in any event-driven data. We will first review some biological and theoretical evidence in neural information processing. We will then present some models for the detection of such motifs in arbitrary raster plots, synthetic, biological or artificial (notably from event-based cameras). In particular, we will discuss models which exploit the variety of synaptic delays on the dendritic tree. Then, we will try to outline some possible strategies for learning these patterns and finally discuss possible perspectives.
}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\vspace{6pt}

@@

\dataavailability{
%This works is made reproducible using the following tools. First the code reproducing all figures is available on \href{https://github.com/SpikeAI/2022-09_UltraFastCat/blob/main/Readme.md}{GitHub}~\citep{Jeremie2022a}.

%\noindent The paper is available a an \href{https://arxiv.org/abs/2205.03635}{arXiv preprint} with links to previous versions ands to the code. 
Find the associated \noindent \href{https://www.zotero.org/groups/4562620/polychronies}{zotero group} used to regroup relevant literature on the subject.

}
